something about variable-byte encoding

### 2020-02-10 gradu/msc_english_contents.tex
.

Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not possible
to decode i-th element in known well-performing compression algorithms like Huffman or Lempel-Ziv, because the position of the element depends on the data 
before it. While VB encoding doesn't natively support direct access, it is achievable with two succinct data structure operations: rank and select.

<explain rank>

<explain select>

<introduce Bri09>

@@

Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
the index of ith 1 bit in the bit array. Both functions work in constant time (citation?) and they require only a few percents of extra space over
the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
version of random access has already been introduced \citep{Bri09}.

Random access with select query is also possible.  By separating the $c$ bit array and $b$ block array, $b$ contains variable-byte integers 
in readable form. Another upside is that functions next(i) and previous(i) are conveniently available. Rank implementation has

- explain how random access is good

\chapter{Rearranged VB codes}
bl

\chapter{Experimental results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@

\chapter{From tex to pdf}

In Linux, run \texttt{pdflatex filename.tex} and \texttt{bibtex filename.tex} repeatedly until no more warnings are shown. This process can be automised using make-command.

>>>>

Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not natively
possible to decode the i-th element in variable-byte compression algorithms, because the position of the element depends on the preceding data. Direct access 
is achievable with supporting data structures. An obvious solution is to store each element's location, but that adds a very large overhead which removes the 
benefit of compression, but for some compression techniques it adds a very large overhead and thus is not applicable.

Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
space over the bit array B (citation?). Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, 
this requires B to be created in addition to the existing data and the length of B usually has to be close to the length of the data.

VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which significantly shrinks the length of B. Also 
the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to 
i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.

\section{DAC via rank}
Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
$A_n$ and then use $rank_1$ to get direct access. For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second 
least significant bits to $A_2$ and the most significant bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. 
If the continuation bit is set at 0 (meaning there are more chunks to this element), getting the correct index for $A_2$ is calculated with $rank_1$($A_1$, i). 
Small values benefit from this reordering, because the first chunk does not need any calculation. This method causes a small overhead on the data, because each 
continuation bit array needs a support data structure for rank.

(TODO: check if actually first introduced or just first with rank)

do rank/select need to be explained better? (or just referenced to)

<introduce Bri09 better?>

@@

 An effective
version of random access has already been introduced \citep{Bri09}.

- explain how random access is good ?

\chapter{DAC with select query}

Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point.

\chapter{Experimental results}
 - comparison to basic implementation + Bri09

 - do we need to support search()?

\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

@@
